An Empirical Analysis of C Preprocessor Use

### 2000-01-18 macros/papers/emp-use-2.tex
Final tweaks.  I'm OK with this now.

  In \pkg{perl}, unknown identifiers make up 31\% of macro definitions,
  compared to 9.0\% in \pkg{mosaic}, which has the the second largest
  proportion of unknown identifiers, and an average of 6.9\%.  In order to
  avoid link-time collisions when the Perl interpreter is embedded in
  another application, \pkg{perl} conditionally redefines over one thousand
  global symbols to add the prefix \verb|Perl_|, as in {\tt \#define Xpv
  Perl_Xpv}.

@@

\item[Side effects]
  The macro body directly changes program state via assignment (of the form
  {\tt =}, {\tt {\em op}=}, {\tt -{}-}, or {\tt ++}).  
  Indirect side
  effects due to function or macro calls were not counted in this category.
  The side effect may be to a
  global variable, a variable local to the macro body, a macro parameter,
  or a local variable in scope at the point of macro invocation.
  Examples of the four varieties of side effects are:
\begin{verbatim}
    #define INIT_FAIL_STACK() do { fail_stack.avail = 0; } while (0)
    #define SWALLOW_LINE(fp) { int c; while ((c = getc(fp)) != '\n' && c != EOF); }
    #define FREE_VAR(var) if (var) free (var); var = NULL
    #define ADD_CMD(x) { if (cmd) specific_limits++; cmd |= (x); }
\end{verbatim}

@@

\item[Pasting]\label{def:pasting}
  The body uses symbol pasting ({\tt \#\#}), which treats its arguments not
  as tokens but as strings, constructing a new token out of their
  concatenation.  After {\tt \#define \verb|_SIZEOF|(x) \verb|sz_|\#\#x},
  the macro invocation {\tt \verb|_SIZEOF|(int)} expands to the 
  identifier {\tt \verb|sz_int|}.  The resulting identifier might appear literally, or
  only as a pasted identifier, at its other uses.  Pasting is often
  abstracted out into a separate macro\,---\,such as {\tt \#define
  \verb|__CONCAT|(x,y) x \#\# y}\,---\,but such pasting macros are used less
  than typical macros defined in our test suite.

@@

The least frequently redefined macros are those categorized as unknown
identifier.  This is partially due to our method of coalescing multiple
definitions:  macro definitions categorized as unknown identifiers are overridden
by definitions with any other categorization (see
Figure~\ref{fig:category-lub}).  Our
analysis included enough library header files to include some
recognizable definition of most common macros.

@@

Common Cpp constructs could be replaced by a special-purpose syntax.  For
instance, declarations or partial declarations could be made explicit
objects, such as by making type modifiers first-class at compile (or, more
likely, preprocess) time; the language or the preprocessor could include a
{\tt dynamic} declaration modifier for dynamic binding (like Lisp's {\tt
special} delcaration for dynamic variables); and
similar support could be provided for repetitive constructs.  Manipulations
of these objects would then be performed through a clearly-specified
interface rather than via string and token concatenation, easing the
understanding burden on the programmer or tool.  Such uses would also be
visible to the compiler and to program checkers such as lint.  The downside
of this approach is the introduction of a new syntax or new library
functions that may not simplify the program text and that cannot cover
every possible case.

@@

Finally, our results are of interest to software engineering researchers
for all of the above reasons and more.  Since this is the first study
of Cpp usage of which we are aware, it was worth performing simply to
determine whether the results were predictable \emph{a priori}. 
Each aspect of our analysis has been surprising and interesting to
some individuals.  Pursuing more highly focused and deeper analyses
would be worthwhile.

>>>>

  In \pkg{perl}, unknown identifiers make up 31\% of macro definitions,
  compared to 9.0\% in \pkg{mosaic}, which has the the second largest
  proportion of unknown identifiers, and an average of 6.9\%.  In order to
  avoid link-time collisions when the Perl interpreter is embedded in
  another application, \pkg{perl} conditionally redefines over one thousand
  global symbols to add the prefix \verb|Perl_|, as in {\tt \#define Xpv
  \verb|Perl_Xpv|}.

@@

\item[Side effects]
  The macro body directly changes program state via assignment (of the form
  {\tt =}, {\tt {\em op}=}, {\tt -{}-}, or {\tt ++}).  
  Indirect side
  effects due to function or macro calls are not counted in this category.
  The side effect may be to a
  global variable, a variable local to the macro body, a macro parameter,
  or a local variable in scope at the point of macro invocation.
  Examples of the four varieties of side effects are:
\begin{verbatim}
    #define INIT_FAIL_STACK() do { fail_stack.avail = 0; } while (0)
    #define SWALLOW_LINE(fp) { int c; while ((c = getc(fp)) != '\n' && c != EOF); }
    #define FREE_VAR(var) if (var) free (var); var = NULL
    #define ADD_CMD(x) { if (cmd) specific_limits++; cmd |= (x); }
\end{verbatim}

@@

\item[Pasting]\label{def:pasting}
  The body uses symbol pasting ({\tt \#\#}), which treats its arguments not
  as tokens but as strings, constructing a new token out of their
  concatenation.  After {\tt \#define \verb|_SIZEOF|(x) \verb|sz_|\#\#x},
  the macro invocation {\tt \verb|_SIZEOF|(int)} expands to the 
  identifier {\tt \verb|sz_int|}.  The resulting identifier might appear literally, or
  only as a pasted identifier, at its other uses.  Pasting is often
  abstracted out into a separate macro\,---\,such as {\tt \#define
  \verb|__CONCAT|(x,y) x \#\# y}\,---\,but such pasting macros are used less
  frequently than average in our test suite.

@@

The least frequently redefined macros are those categorized as unknown
identifier.  This is partially due to our method of coalescing multiple
definitions:  macro definitions categorized as unknown identifier are overridden
by definitions with any other categorization (see
Figure~\ref{fig:category-lub}).  Our
analysis included enough library header files to include some
recognizable definition of most common macros.

@@

Common Cpp constructs could be replaced by a special-purpose syntax.  For
instance, declarations or partial declarations could be made explicit
objects, such as by making type modifiers first-class at compile (or, more
likely, preprocess) time; the language or the preprocessor could include a
{\tt dynamic} declaration modifier for dynamic binding (like the {\tt
special} declaration for dynamic variables in
Lisp~\cite{commonlisp:languagespec}); and
similar support could be provided for repetitive constructs.  Manipulations
of these objects would then be performed through a clearly-specified
interface rather than via string and token concatenation, easing the
understanding burden on the programmer or tool.  Such uses would also be
visible to the compiler and to program checkers such as lint.  The downside
of this approach is the introduction of a new syntax or new library
functions that may not simplify the program text and that cannot cover
every possible case.

@@

Finally, our results are of interest to software engineering researchers
for all of the above reasons and more.  Since this is the first study
of Cpp usage of which we are aware, it was worth performing simply to
determine whether the results were predictable \emph{a priori}. 
Each aspect of our analysis has been surprising and interesting to
some individuals.  Pursuing more highly focused and deeper analyses
along some of these directions could be worthwhile.
